---
title: "NYC helios setup: population, schools and ventilation"
output:
  html_document:
    toc: true
    toc_float: true
    self_contained: true
params:
  model_population: 100000
  results_dir: "results"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 4.5,
  dpi = 150
)

library(helios)
library(dplyr)
library(ggplot2)
library(scales)

nyc_parameters_file <- file.path("inst", "nyc", "nyc_parameters.R")
if (!file.exists(nyc_parameters_file)) {
  nyc_parameters_file <- system.file("nyc", "nyc_parameters.R", package = "helios", mustWork = TRUE)
}
source(nyc_parameters_file)

# Chart palette: first three validated categorical slots and neutral ink/chrome
# from the reference data-visualisation palette (light mode).
ink <- list(
  primary = "#0b0b0b",
  secondary = "#52514e",
  muted = "#898781",
  grid = "#e1e0d9",
  axis = "#c3c2b7",
  surface = "#fcfcfb"
)
series <- c("#2a78d6", "#eb6834", "#1baf7a")

theme_nyc <- function() {
  theme_minimal(base_size = 11) +
    theme(
      plot.background = element_rect(fill = ink$surface, colour = NA),
      panel.background = element_rect(fill = ink$surface, colour = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = ink$grid, linewidth = 0.3),
      axis.line.x = element_line(colour = ink$axis, linewidth = 0.3),
      axis.text = element_text(colour = ink$muted),
      axis.title = element_text(colour = ink$secondary),
      plot.title = element_text(colour = ink$primary, face = "bold"),
      plot.subtitle = element_text(colour = ink$secondary),
      plot.caption = element_text(colour = ink$muted, hjust = 0),
      plot.title.position = "plot",
      plot.caption.position = "plot",
      legend.position = "top",
      legend.justification = "left",
      legend.title = element_blank(),
      legend.text = element_text(colour = ink$secondary),
      strip.text = element_text(colour = ink$primary, hjust = 0, face = "bold")
    )
}

show_table <- function(df, ...) knitr::kable(df, format.args = list(big.mark = ","), ...)

nyc_county_names <- c(
  "36005" = "Bronx",
  "36047" = "Brooklyn",
  "36061" = "Manhattan",
  "36081" = "Queens",
  "36085" = "Staten Island"
)
```

This report summarises the New York City inputs used in helios-nyc: the synthetic
household population, public school sizes, the population helios generates from
them, and the school ventilation assumptions. Charts are static; each has a table
with the underlying values.

# Key numbers

```{r key-numbers}
households <- baseline_household_demographics_usa
household_size <- rowSums(households)
people_by_age <- colSums(households)

key_numbers <- data.frame(
  Measure = c(
    "Households (five boroughs)",
    "People in households",
    "Mean household size",
    "Public schools (2018-19, open with enrollment)",
    "Students enrolled",
    "Students per full-time-equivalent teacher"
  ),
  Value = c(
    comma(nrow(households)),
    comma(sum(household_size)),
    number(mean(household_size), accuracy = 0.01),
    comma(nrow(schools_nyc)),
    comma(sum(schools_nyc$enrollment)),
    number(nyc_school_student_staff_ratio, accuracy = 0.1)
  )
)
show_table(key_numbers)
```

Sources: households from the RTI synthetic population pipeline (2015-2019 American
Community Survey, New York State filtered to the five boroughs, with stochastic
rounding of household counts); schools from the NCES Common Core of Data school
directory for 2018-19.

# Households and age structure

## People by age group

```{r age-groups}
age_df <- data.frame(
  age_group = factor(
    c("Children (0-18)", "Adults (19-69)", "Elderly (70+)"),
    levels = c("Children (0-18)", "Adults (19-69)", "Elderly (70+)")
  ),
  people = as.numeric(people_by_age[c("child", "adult", "elderly")])
)
age_df$share <- age_df$people / sum(age_df$people)

ggplot(age_df, aes(age_group, people)) +
  geom_col(fill = series[1], width = 0.5) +
  geom_text(
    aes(label = paste0(comma(people), " (", percent(share, accuracy = 0.1), ")")),
    vjust = -0.6,
    colour = ink$secondary,
    size = 3.5
  ) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "People by age group",
    subtitle = "NYC synthetic household population",
    x = NULL,
    y = "People"
  ) +
  theme_nyc()

show_table(
  transmute(age_df, `Age group` = age_group, People = people, Share = percent(share, accuracy = 0.1))
)
```

## Household size: model population versus the Census

```{r household-size}
acs_household_size_share <- c(
  "1" = 0.3229, "2" = 0.2845, "3" = 0.1622, "4" = 0.1251,
  "5" = 0.0595, "6" = 0.0251, "7+" = 0.0207
)
size_levels <- names(acs_household_size_share)
model_size_share <- prop.table(table(factor(
  ifelse(household_size >= 7, "7+", as.character(household_size)),
  levels = size_levels
)))

household_size_df <- rbind(
  data.frame(size = size_levels, source = "RTI synthetic population", share = as.numeric(model_size_share)),
  data.frame(size = size_levels, source = "ACS 2015-2019 (table B11016)", share = as.numeric(acs_household_size_share))
)
household_size_df$size <- factor(household_size_df$size, levels = size_levels)
household_size_df$source <- factor(
  household_size_df$source,
  levels = c("RTI synthetic population", "ACS 2015-2019 (table B11016)")
)

ggplot(household_size_df, aes(size, share, fill = source)) +
  geom_col(position = position_dodge2(width = 0.7, padding = 0.1), width = 0.7) +
  scale_fill_manual(values = series[1:2]) +
  scale_y_continuous(labels = label_percent(accuracy = 1), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Share of households by size",
    subtitle = "Large households are slightly under-represented in the synthetic population",
    x = "People in household",
    y = "Share of households"
  ) +
  theme_nyc()

household_size_table <- tidyr::pivot_wider(household_size_df, names_from = source, values_from = share)
household_size_table[-1] <- lapply(household_size_table[-1], percent, accuracy = 0.01)
show_table(rename(household_size_table, `People in household` = size))
```

# Schools

## Students per school

```{r school-sizes}
schools_above_limit <- sum(schools_nyc$enrollment > 3000)

ggplot(filter(schools_nyc, enrollment <= 3000), aes(enrollment)) +
  geom_histogram(binwidth = 100, boundary = 0, fill = series[1], colour = ink$surface, linewidth = 0.6) +
  scale_x_continuous(labels = label_comma(), breaks = seq(0, 3000, 500)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Students per school",
    subtitle = sprintf(
      "%s NYC public schools, 2018-19 (bins of 100 students; %d schools above 3,000 not shown)",
      comma(nrow(schools_nyc)),
      schools_above_limit
    ),
    x = "Students enrolled",
    y = "Schools"
  ) +
  theme_nyc()

show_table(data.frame(
  Statistic = c("Minimum", "25th percentile", "Median", "Mean", "75th percentile", "95th percentile", "Maximum"),
  `Students enrolled` = round(c(
    min(schools_nyc$enrollment),
    quantile(schools_nyc$enrollment, 0.25),
    median(schools_nyc$enrollment),
    mean(schools_nyc$enrollment),
    quantile(schools_nyc$enrollment, 0.75),
    quantile(schools_nyc$enrollment, 0.95),
    max(schools_nyc$enrollment)
  )),
  check.names = FALSE
))
```

## Students by borough

```{r schools-by-borough}
borough_df <- schools_nyc |>
  mutate(borough = nyc_county_names[county_code]) |>
  group_by(borough) |>
  summarise(schools = n(), students = sum(enrollment), .groups = "drop") |>
  arrange(desc(students)) |>
  mutate(borough = factor(borough, levels = borough))

ggplot(borough_df, aes(borough, students)) +
  geom_col(fill = series[1], width = 0.5) +
  geom_text(aes(label = comma(students)), vjust = -0.6, colour = ink$secondary, size = 3.5) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "Students enrolled by borough",
    subtitle = "NYC public schools, 2018-19",
    x = NULL,
    y = "Students"
  ) +
  theme_nyc()

show_table(rename(borough_df, Borough = borough, Schools = schools, Students = students))
```

# Model population

The charts below use a single population generated by helios with the NYC settings
(`r comma(params$model_population)` people). Households and schools are sampled from
the data above. Workplace sizes come from helios's parametric US distribution (an
offset truncated power law), not NYC data. Leisure location sizes come from a
negative binomial distribution.

```{r model-population}
model_n <- params$model_population
model_parameters <- get_parameters(
  overrides = list(
    human_population = model_n,
    number_initial_S = model_n - 10,
    number_initial_E = 10,
    seed = 1
  ),
  archetype = "flu"
)
model_parameters <- apply_nyc_parameters(model_parameters)
model <- create_variables(model_parameters)

location_sizes <- function(variable) {
  categories <- setdiff(variable$get_categories(), "0")
  vapply(categories, function(x) variable$get_size_of(values = x), numeric(1))
}
model_school_sizes <- location_sizes(model$variables_list$school)
model_workplace_sizes <- location_sizes(model$variables_list$workplace)
model_age <- vapply(
  c("child", "adult", "elderly"),
  function(x) model$variables_list$age_class$get_size_of(values = x),
  numeric(1)
)

show_table(data.frame(
  Measure = c(
    "People",
    "Children / adults / elderly",
    "Households",
    "Schools",
    "People assigned to a school (students and staff)",
    "Workplaces",
    "People assigned to a workplace"
  ),
  Value = c(
    comma(model_n),
    paste(comma(model_age), collapse = " / "),
    comma(length(model$variables_list$household$get_categories())),
    comma(length(model_school_sizes)),
    comma(sum(model_school_sizes)),
    comma(length(model_workplace_sizes)),
    comma(sum(model_workplace_sizes))
  )
))
```

## Workplace sizes

```{r workplace-sizes}
ggplot(data.frame(size = model_workplace_sizes), aes(size)) +
  geom_histogram(bins = 30, fill = series[1], colour = ink$surface, linewidth = 0.6) +
  scale_x_log10(labels = label_comma()) +
  scale_y_continuous(labels = label_comma(), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "People per workplace in the model population",
    subtitle = "Log scale; parametric US distribution",
    x = "People in workplace (log scale)",
    y = "Workplaces"
  ) +
  theme_nyc()

show_table(data.frame(
  Statistic = c("Workplaces", "Median size", "Mean size", "95th percentile", "Largest"),
  Value = round(c(
    length(model_workplace_sizes),
    median(model_workplace_sizes),
    mean(model_workplace_sizes),
    quantile(model_workplace_sizes, 0.95),
    max(model_workplace_sizes)
  ))
))
```

## School sizes: model population versus NYC data

In the model, school size includes staff. Large schools are drawn rarely in a
population of this size, so the model's upper tail is shorter than the data's.

```{r school-size-comparison}
quantile_levels <- c(0.05, 0.25, 0.5, 0.75, 0.95)
show_table(data.frame(
  Quantile = percent(quantile_levels),
  `NYC schools (students)` = round(quantile(schools_nyc$enrollment, quantile_levels)),
  `Model schools (students and staff)` = round(quantile(model_school_sizes, quantile_levels)),
  check.names = FALSE
), row.names = FALSE)
```

# School ventilation

Each school's baseline air changes per hour (ACH) is drawn from a normal distribution
truncated at 0. The main analysis uses inputs chosen to reproduce the school-day
classroom ACH of 2.0 +/- 1.3 measured by Batterman et al. (2017, Indoor Air,
doi:10.1111/ina.12384). The sensitivity scenario enters 2.0 and 1.3 directly.

```{r ventilation}
set.seed(1)
ach_draws <- do.call(rbind, lapply(names(nyc_school_ach_scenarios), function(scenario) {
  s <- nyc_school_ach_scenarios[[scenario]]
  data.frame(
    scenario = scenario,
    ach = truncnorm::rtruncnorm(100000, a = 0, b = Inf, mean = s$mean, sd = s$sd)
  )
}))
ach_summary <- ach_draws |>
  group_by(scenario) |>
  summarise(
    realised_mean = mean(ach),
    realised_sd = sd(ach),
    below_half = mean(ach < 0.5),
    p95 = quantile(ach, 0.95),
    .groups = "drop"
  )
scenario_labels <- setNames(
  sprintf(
    "%s (inputs %.2f / %.2f): realised mean %.2f, SD %.2f",
    c(batterman = "Main analysis", batterman_untruncated = "Sensitivity")[ach_summary$scenario],
    vapply(ach_summary$scenario, function(x) nyc_school_ach_scenarios[[x]]$mean, numeric(1)),
    vapply(ach_summary$scenario, function(x) nyc_school_ach_scenarios[[x]]$sd, numeric(1)),
    ach_summary$realised_mean,
    ach_summary$realised_sd
  ),
  ach_summary$scenario
)

ggplot(ach_draws, aes(ach)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 0.25,
    boundary = 0,
    fill = series[1],
    colour = ink$surface,
    linewidth = 0.6
  ) +
  facet_wrap(~scenario, ncol = 1, labeller = labeller(scenario = scenario_labels)) +
  coord_cartesian(xlim = c(0, 8)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Baseline school ACH distribution",
    x = "Air changes per hour",
    y = "Density"
  ) +
  theme_nyc()

show_table(transmute(
  ach_summary,
  Scenario = c(batterman = "Main analysis", batterman_untruncated = "Sensitivity")[scenario],
  `Realised mean` = round(realised_mean, 2),
  `Realised SD` = round(realised_sd, 2),
  `Schools below 0.5 ACH` = percent(below_half, accuracy = 0.1),
  `95th percentile` = round(p95, 2)
))
```

# Prototype simulation results

```{r results-files}
epidemic_file <- file.path(params$results_dir, "nyc_school_epidemic", "nyc_school_epidemic.rds")
endemic_file <- file.path(params$results_dir, "nyc_school_endemic", "endemic_summary.rds")
intervention_labels <- c(
  ashrae_241 = "Schools brought up to ASHRAE 241",
  nyc_current_purifiers = "NYC's current classroom purifiers"
)
```

Interventions: covered schools brought up to the ASHRAE Standard 241 classroom target
(40 cfm of clean air per person, about `r round(ashrae_241_target_ach, 1)` equivalent
air changes per hour in a typical NYC classroom), and, where run, NYC's current
classroom purifiers (about +`r round(nyc_current_purifier_ach, 1)` equivalent air
changes per hour). Transmission rates are the helios flu values, which give R0 about
1.5 in NYC but have not been formally recalibrated.

## Epidemic (single flu season)

```{r epidemic-results, eval = file.exists(epidemic_file), fig.height = 5}
epidemic <- readRDS(epidemic_file)
epidemic_summary <- epidemic$totals |>
  filter(intervention != "none") |>
  group_by(intervention, coverage, coverage_type) |>
  summarise(
    mean_averted = mean(infections_averted_pct),
    se_averted = sd(infections_averted_pct) / sqrt(n()),
    replicates = n(),
    .groups = "drop"
  ) |>
  mutate(
    placement = factor(
      c(random = "Random placement", targeted_riskiness = "Targeted to lowest-ACH schools")[coverage_type],
      levels = c("Random placement", "Targeted to lowest-ACH schools")
    ),
    coverage_label = factor(percent(coverage), levels = percent(sort(unique(coverage))))
  )
epidemic_summary$intervention_label <- factor(
  intervention_labels[epidemic_summary$intervention],
  levels = intervention_labels[intervention_labels %in% intervention_labels[epidemic_summary$intervention]]
)
baseline_attack <- mean(epidemic$totals$attack_rate[epidemic$totals$intervention == "none"])

ggplot(epidemic_summary, aes(coverage_label, mean_averted, fill = placement)) +
  facet_wrap(~intervention_label, ncol = 1) +
  geom_col(position = position_dodge2(width = 0.7, padding = 0.1), width = 0.7) +
  geom_errorbar(
    aes(ymin = mean_averted - se_averted, ymax = mean_averted + se_averted),
    position = position_dodge(width = 0.7),
    width = 0.15,
    colour = ink$secondary,
    linewidth = 0.4
  ) +
  scale_fill_manual(values = series[1:2]) +
  scale_y_continuous(labels = function(x) paste0(x, "%"), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Infections averted by air cleaners in schools",
    subtitle = sprintf(
      "Epidemic, %s people; baseline attack rate %s; bars show mean, whiskers +/- 1 standard error",
      comma(epidemic$settings$population),
      percent(baseline_attack, accuracy = 0.1)
    ),
    x = "Share of schools covered",
    y = "Infections averted"
  ) +
  theme_nyc()

show_table(transmute(
  epidemic_summary,
  Intervention = intervention_label,
  Coverage = coverage_label,
  Placement = placement,
  `Infections averted (%)` = round(mean_averted, 1),
  `Standard error` = round(se_averted, 1),
  Replicates = replicates
))
```

```{r epidemic-missing, eval = !file.exists(epidemic_file), results = "asis"}
cat("Epidemic results not found at `", epidemic_file, "`.\n", sep = "")
```

## Endemic (five years after burn-in)

```{r endemic-results, eval = file.exists(endemic_file)}
endemic <- readRDS(endemic_file)
settings <- endemic$settings
endemic_summary <- endemic$totals |>
  group_by(arm_id, coverage, coverage_type) |>
  summarise(
    annual_infections_per_100k = mean(annual_infections_per_100k),
    mean_averted = mean(infections_averted_pct),
    se_averted = sd(infections_averted_pct) / sqrt(n()),
    replicates = n(),
    .groups = "drop"
  ) |>
  arrange(coverage_type, coverage)

cat(sprintf(
  "Settings: population %s; burn-in %d days; analysis %d days; immunity %d days; about %.1f imported infections per day; %d replicates.\n",
  comma(settings$population), settings$burn_in_days, settings$analysis_days,
  settings$duration_immune_days, settings$imported_infections_per_day, settings$n_reps
))

show_table(transmute(
  endemic_summary,
  Arm = arm_id,
  `Annual infections per 100,000` = round(annual_infections_per_100k),
  `Infections averted (%)` = round(mean_averted, 1),
  `Standard error` = round(se_averted, 1),
  Replicates = replicates
))
```

```{r endemic-missing, eval = !file.exists(endemic_file), results = "asis"}
cat("Endemic results not found at `", endemic_file, "`.\n", sep = "")
```
